---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xpca

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/haziqj/xpca/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/haziqj/xpca/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

An implementation of Principal Components Analysis (PCA) specifically designed for cross-covariance matrices. The `xpca` package provides tools for finding the principal components that maximize the explained variance between two sets of multivariate random variables. 

## Installation

You can install the development version of xpca from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("haziqj/xpca")
```

## Usage

```{r}
library(xpca)
set.seed(123)
x <- matrix(rnorm(100 * 3), nrow = 100, ncol = 3)
y <- matrix(rnorm(100 * 5), nrow = 100, ncol = 5)
xpca(x, y)
```
